---
title: "Affording Fragmented Audiences<br>Multi-Platform Deliberation Within the
Five Star Movement"
subtitle: ""
author: "Francesco Bailo"
institute: "The University of Sydney"
date: "5 April 2024"
output:
  xaringan::moon_reader:
    css: ["default", "assets/sydney-fonts.css", "assets/sydney.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      dev = 'svg', out.width = "45%", fig.width = 6,
                      fig.align="center")

require(fontawesome)

```


---


- This paper is co-authored with Francesco Marolla (Università degli Studi di Milano), Marilù Miotto (Erasmus University Rotterdam) and Giovanni Cassani (Tilburg University). 

- The paper is currently under peer-review. It was accepted for presentation at 2024 Annual Conference of the International Communication Association (Political Communication division). 

---

# Introduction

Briefly introduce the subject matter.

Importance of multi-platform analysis for understanding political movements.

Introduce the Five Star Movement as a case study.

---

## Abstract

- Political movements' online ecosystems extend beyond a single Internet platform.
- Adoption of multiple platforms is driven by the unique user affordances they offer and the organizational outcomes they enable.

### Findings

- **Structural Differences**: Identified within the multi-platform ecosystem of M5S.
- **Visibility and Associability**: Different platforms offer varying degrees of these, affecting organizational outcomes.
  - Meetup.com and the Forum: Higher local visibility and associability.
  - The Blog and Facebook.com: Higher global visibility and associability.
- **Influence on Voting**: Platforms' boundaries influence cross-platform activity and intra-platform communication.
  - Network centrality within Meetup.com and the Forum had a stronger predictive power for the 2012 online primaries results.

---

### Research Questions


---

```{r, echo=FALSE, results='asis'}
xfun::file_string('affordances-outcomes-table.html')
```




