---
title: "Using YouTube Comments to Measure the Geographic Spread of Violence"
subtitle: ""
author: "Francesco Bailo"
date: "7 August 2023"
institute: "The University of Sydney<br><br><br>Soda Labs Alternative Data Workshop @ Monash University"
output:
  xaringan::moon_reader:
    css: ["default", "assets/sydney-fonts.css", "assets/sydney.css"]
    self_contained: false # if true, fonts will be stored locally
    seal: true # show a title slide with YAML information
    includes:
      in_header: "assets/mathjax-equation-numbers.html"
    nature:
      beforeInit: ["assets/remark-zoom.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9' # alternatives '16:9' or '4:3' or others e.g. 13:9
      navigation:
        scroll: false # disable slide transitions by scrolling
---

layout: true
<div style="position: absolute;left:60px;bottom:11px;color:gray;"><small><small><small><a href = 'https://fraba.github.io/presentation/2024-SODA/'>fraba.github.io/presentation/2024-SODA `r icons::fontawesome("link", style = "solid")`</a></small></small></small></div>


```{r setup, include=FALSE, cache=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      dev = 'svg', out.width = "45%", fig.width = 6,
                      fig.align="center")


library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = 'authoryear',
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE,
           no.print.fields = c("urldate","isbn","issn"))
myBib <- ReadBib("assets/biblatex.bib", check = FALSE)
top_icon = function(x) {
  icons::icon_style(
    icons::fontawesome(x),
    position = "fixed", top = 10, right = 10
  )
}
```

---

## Access slides here `r top_icon("link")`

</br></br></br></br></br></br>

.center[.large[[fraba.github.io/presentation/2024-SODA](https://fraba.github.io/presentation/2024-SODA/)]]


</br></br></br></br>

<p style = "font-size: 80px">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&swarr;</p>

---

## Acknowledgement of Country

I would like to acknowledge the Traditional Owners of Australia and  recognise their continuing connection to land, water and culture. The  University of Sydney is located on the land of the Gadigal people  of the Eora Nation. I pay my respects to their Elders, past and present.

---

class: segue-red

# Research problem

---

### Research problem

#### Developing measurements that capture the diffusion of violence across time and space is critical for

- Understanding our societies
  
- Predicting and preventing the onset of physical violence across communities and countries
  
#### Many measures are currently available

- Numerous attempts have been made to develop conflict early warning systems at different scales `r Citep(myBib,c("rod_Review_2024"))`.
  
#### Current limitations

  - Selective measures of event-based violence using social media data have been proposed `r Citep(myBib,c("muchlinski_We_2021"))`.
  
  - Lack of an effective attempt to construct a spatially diffuse measurement of references to violence using everyday social media interactions

---
  
#### Potential of social media data

.content-box-yellow[

  - Can surface violence that is not reported or underreported in traditional news sources (e.g., criminal violence or state violence)
  
  - Accounts for instances of verbal violence
  
  - Volume and constant flow are apt to capture rare events at granular spatiotemporal scales

]  
  
#### Benefits

.content-box-purple[

  - Particularly helpful in capturing the exposure to violence in marginalised and geographically remote communities worldwide
  
]

---

### Background

#### Recent advancements

  - Predictive analytics have significantly influenced forecasting methods for political and social violence.
  
  - `r Citet(myBib,c("mueller_Hard_2022"))`: Employed natural language processing to predict armed conflict using newspaper text data, highlighting the predictive power of media-generated data.
  
#### ViEWS

- `r Citet(myBib,c("hegre_ViEWS_2019"))`: Introduced ViEWS, a political violence early-warning system combining various models.

- Utilizes historical conflict, economic conditions, and institutional stability, along with logistic regression and random forests.

- Provides detailed three-year forecasts for African regions, identifying both persistent and emergent conflict zones.

---

#### State-space approach

- `r Citet(myBib,c("lindholm_Predicting_2022"))`: Explored predictive modelling using a state-space approach to estimate monthly risk and severity of political violence.

- Granular predictions at the grid-cell level offer nuanced understanding of violence probabilities and localized forecasting.

#### Forecasting methods overview

- `r Citet(myBib,c("schrodt_Databased_2013"))`: Broad overview of forecasting methods for armed violence.
- Highlights diverse analytical techniques and the need for innovative computational approaches.
  


#### Pre-conflict interactions

- `r Citet(myBib,c("dorazio_Kickoff_2015"))`: Used data from the Integrated Crisis Early Warning System to identify patterns that forecast intra-state conflict.
- Methodological approach to conflict prediction complementing broader models.

---

#### Predictive model expectations

- `r Citet(myBib,c("cederman_Predicting_2017"))`: Examined realistic scopes for predictive models of armed conflict.
- Suggested short-term forecasts within confined geographical and temporal limits are feasible and useful.

#### Violence early-warning systems

- `r Citet(myBib,c("bazzi_Promise_2022"))`: Evaluated violence early-warning systems using local data from Colombia and Indonesia.
- Found challenges in predicting new outbreaks, highlighting limitations of predictive models.

#### Domestic turmoil and international conflict

- `r Citet(myBib,c("amarasinghe_Diverting_2022"))`: Investigated how governments use foreign policy to divert attention from domestic issues.
- Provides a broader socio-political context to conflict dynamics.

---

.content-box-yellow[

#### Conclusion

- These studies enhance understanding of violence measurement and prediction complexities.
- Highlight the potential of integrating various data sources, including social media, into predictive models for real-time, granular insights crucial for early warning systems and strategic interventions.

]

---

class: segue-red

# Research goal and approach

---

## Research goal

.content-box-yellow[

### a. Developing a highly world-wide granular set of measures of perceived violence, with precise spatial and temporal resolution.

1. Measures will differentiate the **vector** of perceived violence and the **actors** involved

2. Measures will anchor the perceived violence to public **events**.

### b. Fine-tuning the measure against alternative measures to systematic biases of each approach


### c. Understand how to integrate alternative measures for improving real-time sensing and forecasting 

]

---

## Research approach

1. Set a population grid for the area of interest (i.e., country);

--

2. Set a population density threshold for areas to exclude from data collection because too scarcely populated;

--

3. Identifying YouTube videos 

  a. associated with each cell of the grid within the geographic area of interest using the YouTube API; and 
  
  b. published within the timeframe of interest.

4. Collect all comments posted to the YouTube videos identified in 3.

4. Retrieve information from the texts of the comments to estimate of perceived violence. `r emo::ji("construction")`

5. Aggregate this information to compute a set of measures. `r emo::ji("construction")`


---



---

class: segue-red

# Research Approach

---

---

class: segue-red

# Bibliography

---

.small[

```{r, results = "asis", echo=FALSE}
PrintBibliography(myBib)
```

]